TacticToe: Learning to Prove with Tactics

### 2018-02-20 tactictoe_jv.tex
removing todos

\begin{definition} (Goal subsumption)\\
A goal subsumption $\le$ is a partial relation goals.
Assuming we have a way to estimate the number steps required to solve a goal,
a possibly useful subsumption definition could be: \todo{shorter or equal}
\[g_1 \le g_2  \Leftrightarrow_{def} g_1 \mbox{ has a shorter proof than }
g_2\]
By default, we choose a minimal subsumption defined by:
\[g_1 \le g_2  \Leftrightarrow_{def} g_1 \mbox{ is }\alpha\mbox{-equivalent
to } g_2\]
\end{definition}
\todoi{@TG: Which of the two above?}

We can naturally extend a goal subsumption to lists of goals.
\begin{definition} (Goal list subsumption)\\
A goal list subsumption is an partial relation on lists of goals.
Given two lists of goals $l_1$ and $l_2$, we can define it from goal
subsumption $\le$ by:
\[l_1 \le l_2  \Leftrightarrow_{def} \forall g_2 \in l_2.\ \exists g_1 \in l_1.\
g_1 \le g_2\]
\end{definition}

@@

\section{Experimental Evaluation}\label{s:experiments}
\todo{Josef: give the specification}
Each re-proving experiment is performed on a single CPU of a server in Prague
with the exception of the experiments
relying on asynchronous \eprover calls that run on two CPUs.

\subsection{Methodology}
The evaluation imitates the construction of the library: For each theorem only
the previous human proofs are known. These are used as the learning base for
the predictions.
To achieve this scenario we re-prove all theorems during a modified build of
\holfour.
As theorems are proved, their tactical proofs and their statements are
recorded and included in the training examples.
For each theorem we first attempt to run the \tactictoe search with a time
limit of 60 seconds, before processing the original proof script.
In this way, the fairness of the experiments is guaranteed by construction.
Only previously declared \sml
variables (essentially tactics, theorems and simpsets) are accessible.
And for each theorem to be re-proven \tactictoe is only trained on previous
proofs.

\todoi{rephrase to something more coherent}
\paragraph{Datasets: optimization and validation}

All top-level theorems from the standard library are considered with the
exception of 440 hard problems (containing a \texttt{let} construction in their
proof) and 1175 easy problems (build from \texttt{save\_thm} calls).
Therefore, during the full-scale experiments, we evaluate 7164 theorems.
We use every tenth theorem of the first third of the standard library of
them for parameter optimization, which amounts to 273 theorems.
Although the evaluation of each set of parameters on its own is fair,
the selection of the best strategy in Section~\ref{sec:tuning} should also be
considered as a learning process. To ensure the global fairness, the final
experiments in Section~\ref{sec:full_exp} runs the best strategy on the full
dataset which is about 30 times larger.

@@

\paragraph{Reconstruction}
Tactic proofs produced by \tactictoe during this experiment are all verifiable
in \holfour. By the design of the proof search, reconstruction of \tactictoe
proof succeeds, unless one of the tactic modifies the state of \holfour in a
way that changes the behavior of a tactic used in the final proof. Such an
event has not yet occurred.
More details on how a proof is extracted from the search tree is given in
Section~\ref{sec:proofdisplay}.
For \eprover, reconstruction rate varies between 95 and 99 percent, depending
on the reconstruction method \cite{}\todo{CK}.

@@

\paragraph{Reinforcement learning}
All our feature vectors have been learned form human proofs. We now can now
also add tactic-goal pairs that appears in final \tactictoe's proof. To prevent
duplication of effort, orthogonalization of those
tactics is essential to have a beneficial effect.
Since recording and re-proving are intertwined during evaluation, the 
additional data is available for the next proof search.
The hope is that the algorithm will improve faster by learning from its own
discovered proofs than from the human proof
scripts~\cite{DBLP:conf/cade/Urban07}. Side experiments indicate that
this one shot reinforcement learning method increases \tactictoe's success rate 
by half a percent.

@@

We first investigate how \tactictoe and \eprover scale as the search time
grows in Figure~\ref{fig:c1}.
In 10 seconds, \tactictoe solves 90 percent of the problems it can
solve in 60 seconds. The analysis is a bit different for \eprover. Indeed,
we can clearly deduce from the bump at 30 seconds that \eprover is using at
least two strategies and probably more. Strategies are a useful method to
fight the exponential decline in the number of newly proven theorems over
time. Therefore, integrating strategy scheduling in \tactictoe could be
something to be experimented with. Overall, we recommend using \tactictoe and
\eprover with a 10 seconds timeout. It is possible also to run them with a
longer time limit, but we advise to do so in parallel to avoid waiting time.
\todo{its strange to ``advise'' in a paper?}

@@

The total effect on the readability of a proof script is depicted in
Example~\ref{ex:pretty}.
\todo{run without prettification}
\begin{example}\label{ex:pretty}\ \\
Discovered proof script for the theorem \texttt{EVERY\_MAP}
\begin{lstlisting}[language=SMLSmall]
boolLib.REWRITE_TAC [DB.fetch "list" "EVERY_CONJ", DB.fetch "list" "EVERY_MEM",
  DB.fetch "list" "EVERY_EL", ldots ,  combinTheory.o_DEF] THEN
BasicProvers.Induct_on [HolKernel.QUOTE " (*#loc 1 11380*)l"] THENL
  [BasicProvers.SRW_TAC [] [],
   simpLib.ASM_SIMP_TAC (BasicProvers.srw_ss ()) [boolLib.DISJ_IMP_THM,
     DB.fetch "list" "MAP", DB.fetch "list" "CONS_11", boolLib.FORALL_AND_THM]]

\end{lstlisting}
Minimized and prettified proof script:
\begin{lstlisting}[language=SMLSmall]
Induct_on `l` THENL
  [SRW_TAC [] [], ASM_SIMP_TAC (srw_ss ()) [DISJ_IMP_THM, FORALL_AND_THM]]
\end{lstlisting}
\end{example}

\section{Case Study}\todo{comment on the case study}
Since the proofs generated by \tactictoe are meant to be parts of a \holfour development, it interesting to compare their quality with the original human proofs.
The quality of a proof in \holfour can be measured in terms of length, readability, maintainability and verification speed.
We study these properties in three examples taken from our full-scale experiment.
We list the time needed by \holfour to check a proof in parentheses.

@@

\textsf{ML4PG}~\cite{DBLP:journals/corr/abs-1212-3618,DBLP:journals/mics/HerasK14}
groups related proofs using clustering
algorithms. It allows \coq users to inspire themselves from similar proofs and
notice
duplicated proofs. Comparatively, our predictions comes from a much more
detailed description of the target goal.\todo{and can be used as an ATP}
%However, we simply create a single label for each tactic call whereas each of
%its
%arguments is treated independently in \textsf{ML4PG}.
%Our choice is motivated by the k-NN algorithm already used in
%\holyhammer for the a of theorems.
%, which can be easly adapted to the selection of tactics but would be improved
%by a
%version that can learn to predict tactics and their arguments independently.

>>>>

\begin{definition} (Goal subsumption)\\
A goal subsumption $\le$ is a partial relation goals.
Assuming we have a way to estimate the number steps required to solve a goal,
a possibly useful subsumption definition could be:
\[g_1 \le g_2  \Leftrightarrow_{def} g_1 \mbox{ has a proof with a length 
shorter 
or equal to the proof of }
g_2\]
By default, we choose a minimal subsumption defined by:
\[g_1 \le g_2  \Leftrightarrow_{def} g_1 \mbox{ is }\alpha\mbox{-equivalent
to } g_2\]
\end{definition}

We can naturally extend any goal subsumption to a subsumption on
lists of goals.
\begin{definition} (Goal list subsumption)\\
A goal list subsumption is an partial relation on lists of goals.
Given two lists of goals $l_1$ and $l_2$, we can define it from goal
subsumption $\le$ by:
\[l_1 \le l_2  \Leftrightarrow_{def} \forall g_2 \in l_2.\ \exists g_1 \in l_1.\
g_1 \le g_2\]
\end{definition}

@@

\section{Experimental Evaluation}\label{s:experiments}
The execution of \tactictoe's main loop in each re-proving experiment is 
performed on a single CPU. An additional CPU is needed for experiments relying 
on asynchronous \eprover calls.

\subsection{Methodology}
The evaluation imitates the construction of the library: For each theorem only
the previous human proofs are known. These are used as the learning base for
the predictions.
To achieve this scenario we re-prove all theorems during a modified build of
\holfour.
As theorems are proved, their tactical proofs and their statements are
recorded and included in the training examples.
For each theorem we first attempt to run the \tactictoe search with a time
limit of 60 seconds before processing the original proof script.
In this way, the fairness of the experiments is guaranteed by construction:
only previously declared \sml variables (essentially tactics, theorems and 
simpsets) are accessible to \tactictoe.

\paragraph{Datasets: optimization and validation}
All top-level theorems from the standard library are considered with the
exception of 440 hard problems (containing a \texttt{let} construction in their
proof) and 1175 easy problems (build from \texttt{save\_thm} calls).
Therefore, during the full-scale experiments, we evaluate 7164 theorems.
We use every tenth theorem of the first third of the standard library for 
parameter optimization, which amounts to 273 theorems.

Although the evaluation of each set of parameters on its own is fair,
the selection of the best strategy in Section~\ref{sec:tuning} should also be
considered as a learning process. To ensure the global fairness, the final
experiments in Section~\ref{sec:full_exp} runs the best strategy on the full
dataset which is about 30 times larger.

@@

\paragraph{Reconstruction}
Tactic proofs produced by \tactictoe during this experiment are all verifiable
in \holfour. By the design of the proof search, reconstruction of \tactictoe
proof succeeds, unless one of the tactic modifies the state of \holfour in a
way that changes the behavior of a tactic used in the final proof. Such an
event has not yet occurred.
More details on how a proof is extracted from the search tree is given in
Section~\ref{sec:proofdisplay}.
Comparatively, we achieve a reconstruction rate of 95 percent for \eprover,
by calling \metis for 2.0 seconds with the set of theorems necessary for 
in \eprover's proof as argument.

@@

\paragraph{Reinforcement learning}
All our feature vectors have been learned form human proofs. We now can now
also add tactic-goal pairs that appears in final \tactictoe's proof. To prevent
duplication of effort, orthogonalization of those
tactics is essential to have a beneficial effect.
Since recording and re-proving are intertwined during evaluation, the 
additional data is available for the next proof search.
The hope is that the algorithm will improve faster by learning from its own
discovered proofs than from the human proof
scripts~\cite{DBLP:conf/cade/Urban07}. Side experiments show that
this one shot reinforcement learning method increases \tactictoe's success rate 
by less than a percent.

@@

We first investigate how \tactictoe and \eprover scale as the search time
grows in Figure~\ref{fig:c1}.
In 10 seconds, \tactictoe solves 90 percent of the problems it can
solve in 60 seconds. The analysis is a bit different for \eprover. Indeed,
we can clearly deduce from the bump at 30 seconds that \eprover is using at
least two strategies and probably more. Strategies are a useful method to
fight the exponential decline in the number of newly proven theorems over
time. Therefore, integrating strategy scheduling in \tactictoe could be
something to be experimented with.

%Overall, a good compromise is to run \tactictoe and \eprover with a 10 seconds 
%timeout. It is posso run them with a
%longer time limit, but we do so in parallel to avoid waiting time.
%\todo{its strange to ``advise'' in a paper?}

@@

The total effect of these minimization and prettification techniques 
on the readability of a proof script is depicted in
Example~\ref{ex:pretty}.

\begin{example}(Theorem \texttt{EVERY\_MAP})\label{ex:pretty}\\\
Proof script before minimization and prettification:
\begin{lstlisting}[language=SMLSmall]
boolLib.REWRITE_TAC [DB.fetch "list" "EVERY_CONJ", DB.fetch "list" "EVERY_MEM",
  DB.fetch "list" "EVERY_EL", ldots ,  combinTheory.o_DEF] THEN
BasicProvers.Induct_on [HolKernel.QUOTE " (*#loc 1 11380*)l"] THENL
  [BasicProvers.SRW_TAC [] [],
   simpLib.ASM_SIMP_TAC (BasicProvers.srw_ss ()) [boolLib.DISJ_IMP_THM,
     DB.fetch "list" "MAP", DB.fetch "list" "CONS_11", boolLib.FORALL_AND_THM]]

\end{lstlisting}
Proof script after minimization and prettification:
\begin{lstlisting}[language=SMLSmall]
Induct_on `l` THENL
  [SRW_TAC [] [], ASM_SIMP_TAC (srw_ss ()) [DISJ_IMP_THM, FORALL_AND_THM]]
\end{lstlisting}
\end{example}

\section{Case Study}
Since the proofs generated by \tactictoe are meant to be parts of a \holfour development, it interesting to compare their quality with the original human proofs.
The quality of a proof in \holfour can be measured in terms of length, readability, maintainability and verification speed.
We study these properties in three examples taken from our full-scale experiment.
We list the time needed by \holfour to check a proof in parentheses.

@@

\textsf{ML4PG}~\cite{DBLP:journals/corr/abs-1212-3618,DBLP:journals/mics/HerasK14}
groups related proofs using clustering
algorithms. It allows \coq users to inspire themselves from similar proofs and
notice duplicated proofs. Comparatively, our predictions comes from a much more
detailed description of the target goal. \tactictoe can also organize the 
predictions to produce verifiable proofs and is not restricted to user 
advice.
%However, we simply create a single label for each tactic call whereas each of
%its
%arguments is treated independently in \textsf{ML4PG}.
%Our choice is motivated by the k-NN algorithm already used in
%\holyhammer for the a of theorems.
%, which can be easly adapted to the selection of tactics but would be improved
%by a
%version that can learn to predict tactics and their arguments independently.
